---
title: "Simulations for different sampling strategies in case-control studies"
output: html_notebook
author: Wouter van Amsterdam, MD
---

This report reviews some case-control sampling strategies, based on:
* Epidemiology - an introduction, K. Rothman
* Clinical Epidemiology, Grobbee & Hoes
* Lectures and assignments from the Epidemiology Master from Julius Center Utrecht, the Netherlands


```{r setup}
require(magrittr)
require(data.table)
require(ggplot2)
```




```{r}
data(diet, package = "Epi")
setDT(diet)
print(help(diet, package = "Epi"))
head(diet)
```

Use energy group as binary exposure determinant (i.e. exposed or non-exposed), treating high energy as exposed

Generate ID variable, ordered on date of inclusion
```{r}
diet[order(doe), id_ordered:=.I]
diet[order(doe)][1:10, list(id_ordered, doe)]
```


Plot cohort
```{r}
diet %>%
  dplyr::mutate(chd = factor(chd)) %>%
  # dplyr::mutate(chd = factor(chd, levels = c(0,1), labels = c(0, 4))) %>%
  ggplot(aes(x = doe, xend = dox, y = id_ordered, yend = id_ordered,
             col = energy.grp)) + 
  geom_segment() + 
  geom_point(aes(x = dox, shape = chd, alpha = chd), color = "black") + scale_shape_manual(values = c(0, 4)) +
  scale_alpha_manual(values = c(0,1)) +
  labs(x = "Date", y = "Participant", title = "Study population inclusion dates (full cohort)")
```


Calculate cohort based effect measures

```{r}
# out1 <- with(diet,
#   Epi::effx(response = chd, type = "binary",
#             exposure = energy.grp, eff = c("RR")
#             ))

out1 <- diet[, table(energy.grp, chd)] %>%
  epitools::oddsratio()

out2 <- diet[, table(energy.grp, chd)] %>%
  epiR::epi.2by2(method = "cohort.count")

```






