---
title: "Confounding vs effect modification in R"
author: "Wouter van Amsterdam"
---

## Preparation
Load data from car milage for several cars.
```{r}
data("mtcars")
head(mtcars)
knitr::opts_chunk$set(echo=TRUE)
```


## Modeling 

The determinant is binary (transmission automatic vs manual), the outcome continuous (gallons per mile)
First create a crude, unadjusted model without potential effect-modification: milage ~ transmission_type
```{r}
fit0 <- lm(mpg~am, data = mtcars)
summary(fit0)
```

Then a model adjusted for potential confounders

```{r}
fit1 <- lm(mpg~am + wt, data = mtcars)
summary(fit1)
```


As we can see, the effect of transmission om milage is confounded by the weight of the car.

Now add a term of interaction (v-engine vs straight engine, use the function as.factor to treat this as categories instead of as a (continous) numeric variable)


```{r}
fit2 <- lm(mpg~am*as.factor(vs)+wt, data = mtcars)
summary(fit2)
```

The beta for transmission is now reported for each category of the modifier.

Now include an interaction term on the whole formula

```{r}
fit3 <- lm(mpg~(am+wt)*as.factor(vs), data = mtcars)
summary(fit3)
```


In this situation, both the beta's for the determinant (manual vs automatic transmission) as the confounder (weight) get estimates for both levels of the modifier (vertical vs straight engine).



```{r}
fit4 <- lm(mpg~am*disp+wt, data = mtcars)
summary(fit4)
```

