---
title: "Risk ratio and odds ratio"
author: "Wouter van Amsterdam, MD"
output:
  html_document: default
  html_notebook: default
---

```{r setup}
library(data.table)
library(epitools)
library(magrittr)
library(knitr)
```


The purpose of this report is to compare odds ratios and risk ratios

Simulate a cohort study, studying the risk of lung cancer within 20 years for smokers and non-smokers, assuming 
continous exposure (the same amount of sigarettes each day) and no cross-over between groups.
- Domain: adults
- Determinant: smoking
- Outcome: lung cancer within 20 years

Assumptions:
- smoking prevalence = 40%
- risk of lung cancer within 20 years for non-smokers = 5%
- risk ratio for smoking = 10 (which is pretty accurate)


Setup experiment
Conventions:
- Add 0 for unexposed group, 1 for exposed group
```{r cohort, results = 'asis'}
# define variables
n = 10000
prev_exp = .4
risk_ratio_exp = 10
r0 = 0.05
r1 = r0 * risk_ratio_exp

# simulate data
set.seed(2)
df <- data.table(
  patient_id = 1:n,
  smoker = as.logical(rbinom(n = n, size = 1, prob = prev_exp))
)

df[smoker == F, lung_cancer:=as.logical(rbinom(n = .N, size = 1, prob = r0))]
df[smoker == T, lung_cancer:=as.logical(rbinom(n = .N, size = 1, prob = r1))]

# create contingency table
kable(df[, table(smoker, lung_cancer)], caption = "Contingency table for cohort")

# calculate table entries
# n0 = n*(1-prev_exp)
# n1 = n-n0
# o0 = n0 * r0
# o1 = n1 * r1

# create table
# cont_tab <- matrix(c(n0-o0, o0, n1-o1, o1), byrow = T, nrow = 2) %>%
#   as.table(dnn = c("smoking", "lung_cancer"))

# kable(cont_tab, caption = "Contingency table for cohort")

```




Now simulate a case control study in this data, with varying numbers of controls (1 on 1 to 1 on 3)
Conventions:
- n00 = number unexposed, without outcome
- n01 = number unexposed, with outcome
- n10 = number exposed, without outcome
- n11 = number exposed, with outcome

```{r}
case_ids <- df[lung_cancer == T, patient_id]
control_ids <- df[lung_cancer == F, patient_id]

ncases = length(case_ids)
ncontrols = seq(from = 1, to = 3, by = 0.25) * ncases

controls <- sapply(ncontrols, function(x) sample(control_ids, size = x, replace = F))

# new data from subsample of data
```


















